---
title: "MS1: Predictors of Sex-Specific Heat Stress Death"
output: html_document
date: "2025-11-08"
---

# Variables include:
#   - Mortality outcome (Died_InTrt)
#   - Demographic variables (Sex, AgeCategory)
#   - Baseline blood measures (Glucose, Ketone, Hematocrit, Corticosterone, Mitochondrial Copy Number, Telomeres)
#   - Treatments: A (Control), B (Heat-Stress)


# Load Libraries
```{r}

library(dplyr)
library(stringr)
library(tidyr)
library(ggplot2)
library(logistf) 
library(lme4)
library(car)
library(ggeffects)
library(tibble)

```
# Clear memory, set working directory, read in datafile & define plot theme
```{r}

# Clears memory
rm(list=ls(all = TRUE))

#Set working directory (assumes data file is available in same location as code)
current_path = rstudioapi::getActiveDocumentContext()$path 

#Read Data File
AHDB_Exp1_MS1 <- read.csv("AHDB_Exp1_MS1.csv")
str(AHDB_Exp1_MS1)

# Define Plot Theme
plot_theme <- theme_classic(base_size = 18) +
  theme(axis.title.y = element_text(vjust = 1.5), 
        axis.title.x = element_blank()) +
  theme(plot.margin = unit(c(.3, .3, .6, .6), "cm"), 
        line = element_line(linewidth = 1.25)) + 
  theme(axis.line.x = element_line(colour = "black", linewidth = 1.25),
        axis.line.y = element_line(colour = "black", linewidth = 1.25)) + 
  theme(axis.text.x = element_text(angle = 360, hjust = .5, 
                                   margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"), 
                                   color = "black"),
        axis.text.y = element_text(margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"), 
                                   color = "black")) +
  theme(axis.ticks.length = unit(-0.3, "cm"),
        axis.ticks = element_line(color = "black", linewidth = 1.25)) + 
  theme(panel.spacing = unit(2, "lines")) +
  theme(strip.background = element_blank()) +
  #theme(legend.position = "none") + 
  theme(plot.title = element_blank()) 
```
### Data Processing & Filtering 
# This is a post hoc analysis of heat-stress induced mortality, therefore treatment A (control) was removed due to individuals not experiencing the heat treatment


```{r}

#Convert Died in Treatment to factor
AHDB_Exp1_MS1 <- AHDB_Exp1_MS1 %>%
  mutate(
    Died_InTrt = factor(Died_InTrt,
                        levels = c("No", "Yes")))  


# Define columns to convert to numeric 

cols_to_convert <- c("Glucose_1", "Glucose_2", "Glucose_3",
                     "Ketone_1", "Ketone_2", "Ketone_3", "Hematocrit")

# Replace "No" with NA in the specified columns
AHDB_Exp1_MS1[cols_to_convert] <- lapply(AHDB_Exp1_MS1[cols_to_convert], function(x) {
  x[x == "No"] <- NA
  x
})

# Convert these columns to numeric
AHDB_Exp1_MS1[cols_to_convert] <- lapply(AHDB_Exp1_MS1[cols_to_convert], as.numeric)

# Calculate average values for Glucose and Ketones
AHDB_Exp1_MS1$Glucose_Avg <- rowMeans(AHDB_Exp1_MS1[, c("Glucose_1", "Glucose_2", "Glucose_3")],
                                                               na.rm = TRUE)
AHDB_Exp1_MS1$Ketone_Avg <- rowMeans(AHDB_Exp1_MS1[, c("Ketone_1", "Ketone_2", "Ketone_3")],
                                                              na.rm = TRUE)


# Create a Female ZEFI baseline dataset for Hypothesis 2
AHDB_Exp1_MS1_FemaleOnly <- AHDB_Exp1_MS1 %>%
  filter(
    Sex           == "F",
    TimePoint     == "Baseline"
  )

# Create a Younger Female ZEFI baseline dataset Hypothesis 3
AHDB_Baseline_YoungerFemale <- AHDB_Exp1_MS1 %>%
  filter(
    Sex           == "F",
    AgeCategory     == "Younger"
  )

```
### Data Analysis 
## Hypothesis 1 . Females are more likely to die from acute heat (~ 43 – 43.5C for 5 hours) than males.

# Calculate Sample Size by Sex (Does not account for age)

```{r}

AHDB_Exp1_MS1 %>%
  distinct(ID_Band, Sex) %>%
  summarise(
    Total   = n(),
    Males   = sum(Sex == "M"),
    Females = sum(Sex == "F")
  )

```

# Firth Model -> Sex 
# This runs Firth's logistic regression with the penalized likelihood approach)
# Useful for data with reduced sample size and complete separation in variables (DeathInTrt - No Males Died)

```{r}

firth_fit_sex <- logistf(Died_InTrt ~  Sex ,  
                         data = AHDB_Exp1_MS1)
summary(firth_fit_sex)

exp(coef(firth_fit_sex)) 
exp(confint(firth_fit_sex))

```

# Calculate pseudo-R2 (Proportion of variation of independent variable explained by the dependent variable)
```{r}

ll.null <- firth_fit_sex$loglik[1]
ll.proposed <- firth_fit_sex$loglik[2]
pseudoR2 <- (ll.proposed - ll.null) / ll.null
cat("Pseudo R²:", pseudoR2, "\n")

# Result: 5.9%%

```

# Predicted probabilities for each Sex

```{r}
death_data <- data.frame(Sex = c("M", "F"))
predicted_prob_sex <- predict(firth_fit_sex, death_data, type = "response")
prob_table <- data.frame(Sex = death_data$Sex, Predicted_Probability = predicted_prob_sex)
print(prob_table)

```


# Calculate observed death proportions 

```{r}
obs_death <- AHDB_Exp1_MS1 %>%
  mutate(Death = if_else(Died_InTrt == "Yes", 1, 0)) %>%
  group_by(Sex, AgeCategory) %>%
  summarize(
    N = n(),
    DeathCount = sum(Death, na.rm = TRUE),
    Proportion = DeathCount / N
  )
print(obs_death)
```

```{r}

# Dot Plot
ggplot(obs_death, aes(x = AgeCategory , y = Proportion)) +
  geom_point(aes(color = Sex), size = 5) +  # Points
  labs(
    title = "High Mortality of Young Females During Acute Heat",
    x = "Age Category",
    y = "Observed Death Proportions"
  ) +
  theme_classic()

# Bar Graph
# Create a plotting version of the data with a tiny positive value for zero bars
obs_death_plot <- obs_death %>%
  mutate(Proportion_fixed = ifelse(Proportion == 0, 0.001, Proportion))

ggplot(obs_death_plot, aes(x = AgeCategory, y = Proportion_fixed, fill = Sex)) +
  geom_col(position = position_dodge(width = 0.9), width = 0.8) +
  # Apply custom colors for Sex
  scale_fill_manual(values = c( "#D3D3D3", "#000000")) +
  labs(
#    title = "High mortality of younger females during acute heat",
    x = "Age Category",
    y = "Observed Mortality"
  ) +
  theme_classic(base_size = 18)
```

# Firth Model - Sex + Age 
```{r}

firth_fit_sex_age<- logistf(Died_InTrt ~  Sex + AgeCategory, #+ Sex*AgeCategory,  
                            data = AHDB_Exp1_MS1)
summary(firth_fit_sex_age)
#Age & Sex interaction was removed by backward regression 

exp(coef(firth_fit_sex_age)) 
exp(confint(firth_fit_sex_age))

# Sex p= 4.370709e-02 = 0.044*
# Age p= 6.214754e-02 = 0.062
```
# Calculate pseudo-R2 (Proportion of variation of independent variable explained by the dependent variable)

```{r}

ll.null <- firth_fit_sex_age$loglik[1]
ll.proposed <- firth_fit_sex_age$loglik[2]
pseudoR2 <- (ll.proposed - ll.null) / ll.null
cat("Pseudo R²:", pseudoR2, "\n")

# Pseudo R²: 0.1266361
```
# Calculate Predicted Probabilities with both Age & Sex

```{r}

death_data <- data.frame(
  Sex         = c("M","F","F","M"),
  AgeCategory = c("Younger","Older","Younger","Older")
)
predicted_prob_sex_age <- predict(firth_fit_sex_age, death_data, type = "response")
prob_table_sex_age <- data.frame(Sex = death_data$Sex, AgeCategory =death_data$AgeCategory, Predicted_Probability = predicted_prob_sex_age)
print(prob_table_sex_age)

# Results: Younger Females have a 22.8% likelihood of succumbing to acute heat exposure compared to 3.2% for the younger males whereas older females have 6.3% chance of mortality compared to 0.7% for older males. 


```
## Hypothesis 2. Younger adult females are more likely to die from acute heat (~ 43–43.5C for 5 hours) than older young adult females. 

# Firth - Younger Females Only

```{r}
firth_fit_age <- logistf(Died_InTrt ~  AgeCategory, 
                         data = AHDB_Exp1_MS1_FemaleOnly)
summary(firth_fit_age)
exp(coef(firth_fit_age)) 
exp(confint(firth_fit_age))

# Age p= 6.141039e-02 = 0.061
```

# Calculate "Pseudo R²

```{r}

ll.null <- firth_fit_age$loglik[1]
ll.proposed <- firth_fit_age$loglik[2]
pseudoR2 <- (ll.proposed - ll.null) / ll.null
cat("Pseudo R²:", pseudoR2, "\n")

# Pseudo R²: 0.06515706
```

# Predicted probabilities for each AgeCategory

```{r}

deathdata_H2 <- data.frame(AgeCategory = c("Older", "Younger"))
predicted_prob_age <- predict(firth_fit_age, newdata = deathdata_H2, type = "response")
prob_table <- data.frame(Age = deathdata_H2$Age, Predicted_Probability = predicted_prob_age)
print(prob_table)

```

## Hypothesis 3. Among young adult females, there will be metabolic or damage variables in the baseline blood samples (2 weeks prior) may correlated with probability of dying due to acute heat (~43 – 43.5C for 5 hours). 

# Remove rows with missing values
```{r}

Younger_females_Glu <- AHDB_Baseline_YoungerFemale %>%
  filter(!is.na(Glucose_Avg))

```

# Calculate Mean Glucose Values & plot residuals

```{r}

mean(Younger_females_Glu$Glucose_Avg)
sd(Younger_females_Glu$Glucose_Avg)
min(Younger_females_Glu$Glucose_Avg)
max(Younger_females_Glu$Glucose_Avg)
hist(Younger_females_Glu$Glucose_Avg)

```

# Calcualte Sample size For Glucose Model

```{r}

Younger_females_Glu %>%
  distinct(ID_Band, Sex) %>%
  summarise(
    Total   = n(),
    Females = sum(Sex == "F")
  )

```
```{r}

Glucose_Model <- glm(Died_InTrt ~ Glucose_Avg,
                     data = Younger_females_Glu,
                     family = binomial)

summary(Glucose_Model)
```
# Remove rows with missing values & create ketone only subsetted dataset
```{r}

Younger_females_Ket <- AHDB_Baseline_YoungerFemale %>%
  filter(!is.na(Ketone_Avg))

```

# Calculate Mean Ketone Values & plot residual

```{r}

mean(Younger_females_Ket$Ketone_Avg)
sd(Younger_females_Ket$Ketone_Avg)
min(Younger_females_Ket$Ketone_Avg)
max(Younger_females_Ket$Ketone_Avg)
hist(Younger_females_Ket$Ketone_Avg)

```

# Calcualte Sample size For Ketone Model (Includes Lysed Plasma Appearance Individuals)
```{r}

Younger_females_Ket %>%
  distinct(ID_Band, Sex) %>%
  summarise(
    Total   = n(),
    Females = sum(Sex == "F")
  )

# Sample Size: 43
```
# GLMER -> Ketone Model (Includes Lysed Plasma Individuals)

```{r}

Ketone_Model <- glm(Died_InTrt ~ Ketone_Avg,
                      data = Younger_females_Ket,
                      family = binomial)

summary(Ketone_Model)

```
# Remove rows with missing values & create hematocrit dataset
```{r}

Younger_females_Hem <- AHDB_Baseline_YoungerFemale %>%
  filter(!is.na(Hematocrit))

```

# Calculate Mean Hematocrit Values & plot residuals
```{r}

mean(Younger_females_Hem$Hematocrit)
sd(Younger_females_Hem$Hematocrit)
min(Younger_females_Hem$Hematocrit)
max(Younger_females_Hem$Hematocrit)
hist(Younger_females_Hem$Hematocrit)

```
# Calculate Sample Size

```{r}

Younger_females_Hem %>%
  distinct(ID_Band, Sex) %>%
  summarise(
    Total   = n(),
    Females = sum(Sex == "F")
  )

# Sample Size: 40
```
```{r}

# Remove rows with missing values
Younger_females_Hem_NotLysed <- AHDB_Baseline_YoungerFemale %>%
  filter(!is.na(Hematocrit))

# Capture dataset before plasma‐color filtering
Younger_females_Hem_before_plasma <- Younger_females_Hem_NotLysed

# Remove individuals with lysed plasma appearance
Younger_females_Hem_NotLysed <- Younger_females_Hem_NotLysed %>%
  filter(Plasma_Color %in% c("U", "U/L", "Y"))

#Counts how many individuals were removed due to lysis or being yolky
removed_plasma_rows <- Younger_females_Hem_before_plasma %>%
  filter(!Plasma_Color %in% c("U", "U/L", "Y"))

# Total rows (samples) removed
n_removed_rows <- nrow(removed_plasma_rows)
cat("Rows removed due to plasma appearance:", n_removed_rows, "\n")

# Unique individuals removed (by ID_Band)
n_removed_individuals <- removed_plasma_rows %>%
  distinct(ID_Band) %>%
  nrow()
cat("Unique individuals removed:", n_removed_individuals, "\n")
# 10 individuals removed due to plasma appearance

```
# Recalculate Sample Size

```{r}

Younger_females_Hem_NotLysed %>%
  distinct(ID_Band, Sex) %>%
  summarise(
    Total   = n(),
    Females = sum(Sex == "F")
  )

# Sample Size: 30
```
```{r}

Hematocrit_Model <- glm(Died_InTrt ~ Hematocrit,
                        data = Younger_females_Hem_NotLysed,
                        family = binomial)

summary(Hematocrit_Model)
```
# Remove rows with missing values & create CORT dataset
```{r}

Younger_females_CORT <- AHDB_Baseline_YoungerFemale %>%
  filter(!is.na(CORT.ng.mL.))

```

# Calculate Mean CORT Values & plot residuals
```{r}

mean(Younger_females_CORT$CORT.ng.mL.)
sd(Younger_females_CORT$CORT.ng.mL.)
min(Younger_females_CORT$CORT.ng.mL.)
max(Younger_females_CORT$CORT.ng.mL.)
hist(Younger_females_CORT$CORT.ng.mL.)

```
# Calculate Sample Size

```{r}

Younger_females_CORT %>%
  distinct(ID_Band, Sex) %>%
  summarise(
    Total   = n(),
    Females = sum(Sex == "F")
  )

# Sample Size: 40
```
```{r}

CORT_Model <- glm(Died_InTrt ~ CORT.ng.mL.,
                        data = Younger_females_CORT,
                        family = binomial)

summary(CORT_Model)
```
```{r, fig.height=2, fig.width=6}
# Glucose Plot
p1 <- ggplot(Younger_females_Glu, aes(x = Died_InTrt, y = Glucose_Avg)) +
  geom_violin(aes(fill = Died_InTrt), trim = FALSE, alpha = 0.9) +  # Violin by group
  geom_jitter(width = 0.1, alpha = 0.7, size = 1.5, color = "black") +  # Add data points
  scale_fill_manual(values = c("#C0504D", "#153F70")) +  # Custom colors
  labs(
    x = "Died In Treatment?",
    y = "Avg. Glucose"
  ) +
  theme_classic(base_size = 18) +
  theme(legend.position = "none")  # Removes the legend

# Ketone Plot
p2 <- ggplot(Younger_females_Ket, aes(x = Died_InTrt, y = Ketone_Avg)) +
  geom_violin(aes(fill = Died_InTrt), trim = FALSE, alpha = 0.9) +  # Violin by group
  geom_jitter(width = 0.1, alpha = 0.7, size = 1.5, color = "black") +  # Add data points
  scale_fill_manual(values = c("#C0504D", "#153F70")) +  # Custom colors
  labs(
    x = "Died In Treatment?",
    y = "Avg. Ketone"
  ) +
  theme_classic(base_size = 18) +
  theme(legend.position = "none")  # Removes the legend

# Hematocrit Plot
p3 <- ggplot(Younger_females_Hem_NotLysed, aes(x = Died_InTrt, y = Hematocrit)) +
  geom_violin(aes(fill = Died_InTrt), trim = FALSE, alpha = 0.9) +  # Violin by group
  geom_jitter(width = 0.1, alpha = 0.7, size = 1.5, color = "black") +  # Add data points
  scale_fill_manual(values = c("#C0504D", "#153F70")) +  # Custom colors
  labs(
    x = "Died In Treatment?",
    y = "Log-transformed Hematocrit"
  ) +
  theme_classic(base_size = 18) +
  theme(legend.position = "none")  # Removes the legend

library(cowplot)

# Combine into 1 row and 3 columns
combined_blood_measure_plot <- plot_grid(p1, p2, p3, nrow = 1)
combined_blood_measure_plot

#ggsave("Combined_Blood_Plot.png", combined_blood_measure_plot, width = 12, height = 4)

```

# Remove rows with missing values & create Mito dataset
```{r}

Younger_females_Mito <- AHDB_Baseline_YoungerFemale %>%
  filter(!is.na(mtDNA.Mean))

```

# GLMER -> mtDNA Copy Number Model
```{r}

Mito_Model <- glm(Died_InTrt ~  mtDNA.Mean,
                         data = Younger_females_Mito,
                         family = binomial)

summary(Mito_Model)

# Results not significant (p=0.429)

# Plot
ggplot(Younger_females_Mito, aes(x = Died_InTrt, y = mtDNA.Mean)) +
  geom_violin(aes(fill = Died_InTrt), trim = FALSE, alpha = 0.6) +  # Violin by group
  geom_jitter(width = 0.1, alpha = 0.7, size = 1.5, color = "black") +  # Add data points
  labs(
    x = "Died In Treatment?",
    y = "Avg. mtDNA Copy Number"
  ) +
  theme_classic()

```
# Remove rows with missing values & create Telomeres dataset
```{r}

Younger_females_Telo <- AHDB_Baseline_YoungerFemale %>%
  filter(!is.na(Telomeres.per.cell))

```

# GLMER -> Telomere (T/S ratio) Model (WITH LYSED BUT NO Mom_ID)
```{r}

Telomeres_Model <- glm(Died_InTrt ~  Telomeres.per.cell,
                         data = Younger_females_Telo,
                         family = binomial)

summary(Telomeres_Model)

# Results not significant (p=0.942)

# Plot
ggplot(Younger_females_Telo, aes(x = Died_InTrt, y = Telomeres.per.cell)) +
  geom_violin(aes(fill = Died_InTrt), trim = FALSE, alpha = 0.6) +  # Violin by group
  geom_jitter(width = 0.1, alpha = 0.7, size = 1.5, color = "black") +  # Add data points
  labs(
    x = "Died In Treatment?",
    y = "Telomeres (T/S ratio)"
  ) +
  theme_classic()

```

# Double checking SCNAG
```{r}

# Remove rows with missing values & create SCNAG dataset


Younger_females_SCNAG <- AHDB_Baseline_YoungerFemale %>%
  filter(!is.na(Cq.Mean_SCNAG))


SCNAG <- glm(Died_InTrt ~  Cq.Mean_SCNAG,
                         data = Younger_females_SCNAG,
                         family = binomial)

summary(SCNAG)

# Results not significant (p=0.848)

# Plot
ggplot(Younger_females_SCNAG, aes(x = Died_InTrt, y = Cq.Mean_SCNAG)) +
  geom_violin(aes(fill = Died_InTrt), trim = FALSE, alpha = 0.6) +  # Violin by group
  geom_jitter(width = 0.1, alpha = 0.7, size = 1.5, color = "black") +  # Add data points
  labs(
    x = "Died In Treatment?",
    y = "Cq.Mean_SCNAG"
  ) +
  theme_classic()

```
#Export Model Stats (For all Models)

```{r}

models <- list(
  Glucose_Model  = Glucose_Model,
  Ketone_Model   = Ketone_Model,
  Hematocrit_Model = Hematocrit_Model,
  Telomeres_Model = Telomeres_Model,
  Mito_Model = Mito_Model,
  CORT_Model = CORT_Model
)

extract_model_info <- function(model, model_name) {
  sm <- summary(model)
  coefs <- as.data.frame(sm$coefficients)
  
  # Faster CIs for GLMs (avoids profiling)
  ci <- confint.default(model)
  
  out <- cbind(
    Model = model_name,
    Predictor = rownames(coefs),
    Estimate = coefs[,1],
    Std_Error = coefs[,2],
    Test_Stat = coefs[,3],
    P_Value = coefs[,4],
    CI_Lower = ci[,1],
    CI_Upper = ci[,2]
  )
  
  rownames(out) <- NULL
  return(as.data.frame(out))
}

Model_Results <- bind_rows(
  lapply(names(models), function(nm) extract_model_info(models[[nm]], nm))
)

Model_Results

write.csv(Model_Results, "Model_Summary_Table.csv", row.names = FALSE)

```

