---
title: "AHDB1_Death_TS"
author: "Tonia S. Schwartz"
date: "2025-01-22"
output: 
  md_document:
    variant: gfm
  pdf_document:
  word_document:
  html_document:
    toc: true
    toc_float: true
---

https://github.com/StatQuest/logistic_regression_demo/blob/master/logistic_regression_demo.R
###Read in main data files
```{r, echo=T, message=F, warning=F, results='hide'}
library(multcomp)
library(ggplot2)
library(nlme)
library(grid)
library(gridExtra)
library(emmeans)
library(cowplot)
library(ggplot2)

```


# Read in the data
```{r}
data=read.csv("AHDB_MasterDataSheet.csv")
data
xtabs(~ Died_InTrt + Sex, data=data) # Summarize the number of deaths by Sex
xtabs(~ Died_InTrt + AgeCategory, data=data) # Summarize the number of deaths by Age
```

# Run generalized linear model for Sex
```{r}
logistic <- glm(Died_InTrt ~ Sex, data=data, family="binomial")
summary(logistic)

female.log.odds <- log(9 / 36)
female.log.odds

male.log.odds <- log(0 / 18)
male.log.odds
```

# Now calculate the overall "Pseudo R-squared" and its p-value for Sex
```{r}
## NOTE: Since we are doing logistic regression...
## Null devaince = 2*(0 - LogLikelihood(null model))
##               = -2*LogLikihood(null model)
## Residual deviacne = 2*(0 - LogLikelihood(proposed model))
##                   = -2*LogLikelihood(proposed model)
ll.null <- logistic$null.deviance/-2
ll.proposed <- logistic$deviance/-2
 
## McFadden's Pseudo R^2 = [ LL(Null) - LL(Proposed) ] / LL(Null)
(ll.null - ll.proposed) / ll.null
 
## chi-square value = 2*(LL(Proposed) - LL(Null))
## p-value = 1 - pchisq(chi-square value, df = 2-1)
1 - pchisq(2*(ll.proposed - ll.null), df=1)
1 - pchisq((logistic$null.deviance - logistic$deviance), df=1)
```

## Lastly, let's  see what this logistic regression predicts, given that an individual is either female or male (and no other data about them).
```{r}
predicted.data_sex <- data.frame(
  probability.of.Data=logistic$fitted.values,
  sex=data$Sex)
 
## We can plot the data...
ggplot(data=predicted.data_sex, aes(x=sex, y=probability.of.Data)) +
  geom_point(aes(color=sex), size=5) +
  xlab("Sex") +
  ylab("Predicted probability of Death in Acute Heat in all birds") +
  theme_classic()
 
## Since there are only two probabilities (one for females and one for males),
## we can use a table to summarize the predicted probabilities.
xtabs(~ probability.of.Data + sex, data=predicted.data_sex)
```


# Run a generalized linear model for Age
```{r}
logistic2 <- glm(Died_InTrt ~ AgeCategory, data=data, family="binomial")
summary(logistic2)
```

# Now calculate the overall "Pseudo R-squared" and its p-value for Age
```{r}
## NOTE: Since we are doing logistic regression...
## Null devaince = 2*(0 - LogLikelihood(null model))
##               = -2*LogLikihood(null model)
## Residual deviacne = 2*(0 - LogLikelihood(proposed model))
##                   = -2*LogLikelihood(proposed model)
ll.null <- logistic2$null.deviance/-2
ll.proposed <- logistic2$deviance/-2
 
## McFadden's Pseudo R^2 = [ LL(Null) - LL(Proposed) ] / LL(Null)
(ll.null - ll.proposed) / ll.null
 
## chi-square value = 2*(LL(Proposed) - LL(Null))
## p-value = 1 - pchisq(chi-square value, df = 2-1)
1 - pchisq(2*(ll.proposed - ll.null), df=1)
1 - pchisq((logistic2$null.deviance - logistic2$deviance), df=1)
```

# Lastly, let's  see what this logistic regression predicts, given that an individual is either younger or older (and no other data about them).
```{r}
predicted.data_Age <- data.frame(
  probability.of.Data=logistic2$fitted.values,
  AgeCategory=data$AgeCategory)
 
## We can plot the data...
ggplot(data=predicted.data_Age, aes(x=AgeCategory, y=probability.of.Data)) +
  geom_point(aes(color=AgeCategory), size=5) +
  xlab("Age") +
  ylab("Predicted probability of Death in Acute Heat in all birds") +
  theme_classic()
 
## Since there are only two probabilities (one for females and one for males),
## we can use a table to summarize the predicted probabilities.
xtabs(~ probability.of.Data + AgeCategory, data=predicted.data_Age)
```




# Run a generalized linear model for Sex and Age
```{r}
logistic_Age_Sex <- glm(Died_InTrt ~ AgeCategory + Sex, data=data, family="binomial")
summary(logistic_Age_Sex)
```

# Now calculate the overall "Pseudo R-squared" and its p-value for sex and age
```{r}
## NOTE: Since we are doing logistic regression...
## Null devaince = 2*(0 - LogLikelihood(null model))
##               = -2*LogLikihood(null model)
## Residual deviacne = 2*(0 - LogLikelihood(proposed model))
##                   = -2*LogLikelihood(proposed model)
ll.null <- logistic_Age_Sex$null.deviance/-2
ll.proposed <- logistic_Age_Sex$deviance/-2
 
## McFadden's Pseudo R^2 = [ LL(Null) - LL(Proposed) ] / LL(Null)
(ll.null - ll.proposed) / ll.null
 
## chi-square value = 2*(LL(Proposed) - LL(Null))
## p-value = 1 - pchisq(chi-square value, df = 2-1)
1 - pchisq(2*(ll.proposed - ll.null), df=1)
1 - pchisq((logistic_Age_Sex$null.deviance - logistic_Age_Sex$deviance), df=1)
```

## Lastly, let's  see what this logistic regression predicts, given that an individual is either younger female, older female, younger male, or older male (and no other data about them).
```{r} 
predicted.data_Age_Sex <- data.frame(
  probability.of.Data=logistic_Age_Sex$fitted.values,
  AgeCategory=data$AgeCategory, Sex=data$Sex)
 
## We can plot the data...
ggplot(data=predicted.data_Age_Sex, aes(x=AgeCategory, y=probability.of.Data, fill=Sex)) +
  geom_point(aes(color=Sex), size=5) +
  xlab("Age") +
  ylab("Predicted probability of Death in Acute Heat in all birds") +
  theme_classic()
 
## Since there are only two probabilities (one for females and one for males),
## we can use a table to summarize the predicted probabilities.
xtabs(~ probability.of.Data + AgeCategory + Sex, data=predicted.data_Age_Sex)
```


# Read in the data - Younger animals only
```{r}
YoungerData=read.csv("Younger.csv")
xtabs(~ Died_InTrt + Sex, data=YoungerData)
```

# Run a generalized linear model for sex in the dataframe only containing younger individuals
```{r}
logistic3 <- glm(Died_InTrt ~ Sex, data=YoungerData, family="binomial")
summary(logistic3)

Youngfemale.log.odds <- log(9 / 36)
Youngfemale.log.odds

Youngmale.log.odds <- log(0 / 18)
Youngmale.log.odds
```

# Now calculate the overall "Pseudo R-squared" and its p-value for younger individuals by sex
```{r}
 ## NOTE: Since we are doing logistic regression...
## Null devaince = 2*(0 - LogLikelihood(null model))
##               = -2*LogLikihood(null model)
## Residual deviacne = 2*(0 - LogLikelihood(proposed model))
##                   = -2*LogLikelihood(proposed model)
ll.null <- logistic3$null.deviance/-2
ll.proposed <- logistic3$deviance/-2
 
## McFadden's Pseudo R^2 = [ LL(Null) - LL(Proposed) ] / LL(Null)
(ll.null - ll.proposed) / ll.null
 
## chi-square value = 2*(LL(Proposed) - LL(Null))
## p-value = 1 - pchisq(chi-square value, df = 2-1)
1 - pchisq(2*(ll.proposed - ll.null), df=1)
1 - pchisq((logistic3$null.deviance - logistic3$deviance), df=1)
```

# Lastly, let's  see what this logistic regression predicts.
```{r}
predicted.data <- data.frame(
  probability.of.YoungerData=logistic3$fitted.values,
  sex=YoungerData$Sex)
 
## We can plot the data...
ggplot(data=predicted.data, aes(x=sex, y=probability.of.YoungerData)) +
  geom_point(aes(color=sex), size=5) +
  xlab("Sex") +
  ylab("Predicted Probability of Acute Heat Death in Younger Birds") +
  theme_classic()
 
## Since there are only two probabilities (one for females and one for males),
## we can use a table to summarize the predicted probabilities.
xtabs(~ probability.of.YoungerData + sex, data=predicted.data)
```
 
 