---
title: "AHDB1 Manuscript 1 (Sex-specifc Mortality) Redo (Heidinger)"
author: "Brynleigh Payne"
date: "2025-08-04"
output: 
  md_document:
    variant: gfm
  pdf_document:
  word_document:
  html_document:
    toc: true
    toc_float: true
---
### Prepare the data for analysis

# Import qPCR output for each plate 1) the Single Copy Autosomal Gene (EEF2) and mtDNA gene multiplex and 2) the Telomere reaction. 
```{r}
SCNAG <- read.csv("AHDB1_Manuscript1_Deaths_Multiplex_2025-08-01 13-38-10_795BR20744 -  Quantification Cq Results.csv")
dim(SCNAG)

Telo <- read.csv("AHDB1_Manuscript1_Deaths_Telomere_Heidinger_2025-08-01 11-52-51_795BR20744 -  Quantification Cq Results.csv")
dim(Telo)
```

# Edit 
```{r}
# Concatenate data across runs for the same samples
Plate1 <- rbind(SCNAG, Telo)
dim(Plate1)

# Add a column called PlateID and fill in correct Plate number to use as a variable in the statistics
Plate1$PlateID <- "Plate1"

# CHECK AND EDIT FOR YOUR DATA. 
#Name Correct Targets based on the fluorphores used in your reaction.
Plate1$Target[Plate1$Fluor == "VIC"] <- "scnag"
Plate1$Target[Plate1$Fluor == "FAM"] <- "mtdna"
Plate1$Target[Plate1$Fluor == "SYBR"] <- "telomeres"
```

# Identify and remove outliers (across the three replicates)
```{r}
# Add a column called "Diff_AVG-Cq"
Plate1$Diff_AVG_Cq <- abs(Plate1$Cq - Plate1$Cq.Mean)

## Add a column called "Flag_outlier"
Plate1$Flag_outlier <- ifelse(Plate1$Diff_AVG_Cq > 0.4001, "yes", "no")
HighCq <- Plate1[Plate1$Diff_AVG_Cq > "0.3001", ]
VeryHighCq <- Plate1[Plate1$Diff_AVG_Cq > "0.4001", ]

# Report number of rows to be removed
print(paste("Number of rows with High Cq:", nrow(VeryHighCq)))
# Make a table called  of the high cq values
HighCq_Samples<- HighCq[c("Well", "Sample", "Fluor", "Diff_AVG_Cq" )]
print(HighCq_Samples)


# Filter and remove rows where Diff_AVG_Cq > 0.4001 for each SampleID
# USed to be max(AVG) but error said that col did not exist
dim(Plate1)
Plate1 <- Plate1 %>%
  group_by(Sample, Target) %>%
  filter(!(Diff_AVG_Cq > 0.4001 & Diff_AVG_Cq == max(Diff_AVG_Cq)))
dim(Plate1)

```


# Additional identification and removal of outliers
```{r}
# Recalculate Cq Mean and Add a column called "Cq Mean2"
Plate1 <- Plate1 %>%
  group_by(Sample, Target) %>%
  mutate(
    Cq.Mean2 = mean(Cq),
    SQ.Mean2 = mean(Starting.Quantity..SQ.)) %>%
  ungroup()
dim(Plate1)

# Add a column called "Diff_AVG-Cq_2" that contains the difference between new Cq and new mean
Plate1$Diff_AVG_Cq_2 <- abs(Plate1$Cq - Plate1$Cq.Mean2)

# Add a column called "Flag_outlier_2"
Plate1$Flag_outlier_2 <- ifelse(Plate1$Diff_AVG_Cq_2 > 0.401, "yes", "no")
VeryHighCq_2 <- Plate1[Plate1$Diff_AVG_Cq > "0.401", ]
# Report number of rows to be removed
print(paste("Number of rows with High Cq:", nrow(VeryHighCq_2)))
# Make a table called  of the high cq values
HighCq_2_Samples<- HighCq[c("Well", "Sample", "Fluor", "Diff_AVG_Cq" )]
print(HighCq_2_Samples)

# Remove the rows with a high Cq Outliers
# USed to be max(AVG) but error said that col did not exist

Plate1 <- Plate1 %>%
  group_by(Sample, Target) %>%
  filter(!(Diff_AVG_Cq_2 > 0.4001 & Diff_AVG_Cq_2 == max(Diff_AVG_Cq_2)))

dim(Plate1)

###############  If any "Samples" do not have more than two rows (replicates left), remove
Plate1 <- Plate1 %>%
  group_by(Sample, Target) %>%
  filter(n() >= 2) %>%
  ungroup()
dim(Plate1)

```


# Remove negative controls and standards
```{r}
# rows that have "NEG", "POS" in column "Sample" and remove rows with "STD" in Sample "Content"
Plate1 <- Plate1 %>%
  filter(!str_detect(Content, "Std-*")) %>%
  filter(!str_detect(Content, "NTC")) %>%
  filter(!str_detect(Sample, "Neg")) %>%
  filter(!str_detect(Sample, "STD*"))
dim(Plate1)

```

# Subset dataset based on the value in "Target" column
```{r}
unique_targets <- unique(Plate1$Target)
# Create a list to store the subset dataframes
subset_dfs <- list()
# Loop through each unique value in 'Target', subset the dataframe, and store in subset_dfs
for (target_value in unique_targets) {
  subset_df <- subset(Plate1, Target == target_value)
  subset_dfs[[target_value]] <- subset_df
}

```

# Now subset_dfs is a list where each element is a dataframe containing rows for each unique 'Target' value
```{r}
Plate1_SCNAG<-print(subset_dfs[["scnag"]])
Plate1_SCNAG<-Plate1_SCNAG[ ,c("PlateID", "Well", "Sample", "Target", "Cq", "Cq.Mean",  "Flag_outlier", "SQ.Mean")]
Plate1_SCNAG <- Plate1_SCNAG %>% 
  rename(Target_SCNAG = Target, Cq_SCNAG = Cq, Cq.Mean_SCNAG = Cq.Mean, Flag_outlier_SCNAG=Flag_outlier, SQ.Mean_SCNAG=SQ.Mean)

Plate1_mtDNA<-print(subset_dfs[["mtdna"]])
Plate1_mtDNA<-Plate1_mtDNA[ ,c("PlateID", "Well", "Sample", "Target", "Cq", "Cq.Mean", "Flag_outlier", "SQ.Mean")]
Plate1_mtDNA <- Plate1_mtDNA %>% 
  rename(Target_mtDNA = Target, Cq_mtDNA = Cq, Cq.Mean_mtDNA = Cq.Mean, Flag_outlier_mtDNA = Flag_outlier, SQ.Mean_mtDNA = SQ.Mean)

Plate1_Telomeres<-print(subset_dfs[["telomeres"]])
Plate1_Telomeres<-Plate1_Telomeres[ ,c("PlateID", "Well", "Sample", "Target", "Cq", "Cq.Mean", "Flag_outlier", "SQ.Mean")]
Plate1_Telomeres <- Plate1_Telomeres %>% 
  rename(Cq_Telomeres = Cq, Cq.Mean_Telomeres = Cq.Mean, Flag_outlier_Telomeres = Flag_outlier, SQ.Mean_Telomeres = SQ.Mean)

```

#  Make a final MPX dataset for by merging the Target datasets horizontally, in rows.
```{r}
Plate1_FinalMPX <- merge(Plate1_SCNAG, Plate1_mtDNA, by = c("PlateID", "Well", "Sample"))

# Normalize mtDNA
# Add a column called mtDNA, and calculate the normalized value 
Plate1_FinalMPX$mtDNA <- (Plate1_FinalMPX$SQ.Mean_mtDNA / Plate1_FinalMPX$SQ.Mean_SCNAG)

# Recalculate mean across the replicates
Plate1_FinalMPX <- Plate1_FinalMPX %>%
  group_by(Sample) %>%
  mutate(
    mtDNA.Mean = mean(mtDNA)) %>%
  ungroup()

```


# Merge final MPX with Telomeres
```{r}
## Reduce datasets to single row per individual containing only the columns we want.
Plate1_FinalMPX <- distinct(Plate1_FinalMPX, PlateID, Sample, SQ.Mean_SCNAG, Cq.Mean_SCNAG, SQ.Mean_mtDNA, mtDNA.Mean)
Plate1_FinalTelo <- distinct(Plate1_Telomeres, PlateID, Sample, SQ.Mean_Telomeres, Cq.Mean_Telomeres)

## Merge the files horizontally
Plate1_FinalData <- merge(Plate1_FinalMPX, Plate1_FinalTelo, by = c("PlateID", "Sample"))

# Normalize Telomeres
Plate1_FinalData <- Plate1_FinalData %>% mutate(Telomeres.per.cell = SQ.Mean_Telomeres / SQ.Mean_SCNAG)

```


# Aggregate to get one row per sample, taking mean of normalized mtDNA and telomeres
```{r}
Plate1_FinalData <- Plate1_FinalData %>%
  group_by(Sample) %>%
  summarize(
    SQ.Mean_SCNAG = mean(SQ.Mean_SCNAG),
    SQ.Mean_mtDNA = mean(SQ.Mean_mtDNA),
    mtDNA.Mean = mean(mtDNA.Mean),
    Cq.Mean_SCNAG = mean(Cq.Mean_SCNAG),
    SQ.Mean_Telomeres = mean(SQ.Mean_Telomeres),
    Cq.Mean_Telomeres = mean(Cq.Mean_Telomeres),
    Telomeres.per.cell = mean(Telomeres.per.cell)
  ) %>%
  ungroup()

```


# Merge Final Data with Trait MetaData for your individuals
```{r}
# Load in Data
Trait <- read.csv("Trait_MetaData.csv")
dim (Trait)

# Merge both datasets
FinalData <- merge(Plate1_FinalData, Trait, by = c("Sample"))

```

# Write the final data file for this plate
```{r}
write.csv(file = "AHDB1_Manuscript1_ESEBqPCR_FinalData.csv", FinalData, row.names = FALSE)

```


### Data analysis

```{r setup, include=FALSE}
library(lme4)
library(lmerTest)
```

# Load data
```{r}
datum <- read.csv("AHDB1_Manuscript1_ESEBqPCR_FinalData.csv")
head(datum)
```

# Convert variables to factors
```{r}
datum$Sample <- as.factor(datum$Sample)
datum$Died_InTrt <- as.factor(datum$Died_InTrt)

str(datum)
```

# Hypothesis 3. Among young adult females, there will be metabolic or damage variables in the baseline blood samples (2 weeks prior) may correlated with probability of dying due to acute heat (~43 – 43.5C for 5 hours).

# GLMER -> mtDNA Copy Number Model (WITH LYSED BUT NO Mom_ID)
```{r}

glmer_model_Mito <- glmer(Died_InTrt ~  mtDNA.Mean +
                            (1 | Cohort),
                         data = datum,
                         family = binomial)

summary(glmer_model_Mito)

# Results not significant (p=0.429)

# Plot
ggplot(datum, aes(x = Died_InTrt, y = mtDNA.Mean)) +
  geom_violin(aes(fill = Died_InTrt), trim = FALSE, alpha = 0.6) +  # Violin by group
  geom_jitter(width = 0.1, alpha = 0.7, size = 1.5, color = "black") +  # Add data points
  labs(
    x = "Died In Treatment?",
    y = "Avg. mtDNA Copy Number"
  ) +
  theme_classic()

```

# GLMER -> Telomere (T/S ratio) Model (WITH LYSED BUT NO Mom_ID)
```{r}

glmer_model_Telo <- glmer(Died_InTrt ~  Telomeres.per.cell +
                            (1 | Cohort),
                         data = datum,
                         family = binomial)

summary(glmer_model_Telo)

# Results not significant (p=0.942)

# Plot
ggplot(datum, aes(x = Died_InTrt, y = Telomeres.per.cell)) +
  geom_violin(aes(fill = Died_InTrt), trim = FALSE, alpha = 0.6) +  # Violin by group
  geom_jitter(width = 0.1, alpha = 0.7, size = 1.5, color = "black") +  # Add data points
  labs(
    x = "Died In Treatment?",
    y = "Telomeres (T/S ratio)"
  ) +
  theme_classic()

```

# Double checking SCNAG
```{r}
glmer_model_SCNAG <- glmer(Died_InTrt ~  Cq.Mean_SCNAG +
                            (1 | Cohort),
                         data = datum,
                         family = binomial)

summary(glmer_model_SCNAG)

# Results not significant (p=0.848)

# Plot
ggplot(datum, aes(x = Died_InTrt, y = Cq.Mean_SCNAG)) +
  geom_violin(aes(fill = Died_InTrt), trim = FALSE, alpha = 0.6) +  # Violin by group
  geom_jitter(width = 0.1, alpha = 0.7, size = 1.5, color = "black") +  # Add data points
  labs(
    x = "Died In Treatment?",
    y = "Cq.Mean_SCNAG"
  ) +
  theme_classic()

```

## ESEB Poster Brynleigh -Graph of Glucose, Ketone, Hematocrit averages

# Models do not include mom ID due to only two being the same within the died in treatment individuals
# Includes cohort as a random effect
```{r, fig.height=2, fig.width=5}
# Mitochondrial Plot
p1 <- ggplot(datum, aes(x = Died_InTrt, y = mtDNA.Mean)) +
  geom_violin(aes(fill = Died_InTrt), trim = FALSE, alpha = 0.9) +  # Violin by group
  geom_jitter(width = 0.1, alpha = 0.7, size = 1.5, color = "black") +  # Add data points
  scale_fill_manual(values = c("#C0504D", "#153F70")) +  # Custom colors
  labs(
    x = "Died In Treatment?",
    y = "Avg. mtDNA Copy Number"
  ) +
  theme_classic(base_size = 18) +
  theme()  # Removes the legend

# Telomere Plot
p2 <- ggplot(datum, aes(x = Died_InTrt, y = Telomeres.per.cell)) +
  geom_violin(aes(fill = Died_InTrt), trim = FALSE, alpha = 0.9) +  # Violin by group
  geom_jitter(width = 0.1, alpha = 0.7, size = 1.5, color = "black") +  # Add data points
  scale_fill_manual(values = c("#C0504D", "#153F70")) +  # Custom colors
  labs(
    x = "Died In Treatment?",
    y = "Telomere Length (T/S Ratio)"
  ) +
  theme_classic(base_size = 18) +
  theme()  # Removes the legend

library(cowplot)

# Combine into 1 row and 3 columns
combined_qPCR_plot <- plot_grid(p1, p2, nrow = 1)
combined_qPCR_plot

#ggsave("Combined_qPCR_Plot.png", combined_qPCR_plot, width = 12, height = 4)

```





