---
title: "MS1: Predictors of Sex-Specific Heat Stress Death"
output: html_notebook
---

# Variables include:
#   - Mortality outcome (Died_InTrt)
#   - Demographic variables (Sex, AgeCategory)
#   - Baseline blood measures (Glucose, Ketone, Hematocrit)
#   - Treatments: A (Control), B (Heat-Stress)
---

# Load Libraries
```{r}

library(dplyr)
library(stringr)
library(tidyr)
library(ggplot2)
library(logistf) 
library(lme4)
library(car)
library(ggeffects)
library(tibble)

```

# Clear memory, set working directory, read in datafile & define plot theme
```{r}

# Clears memory
rm(list=ls(all = TRUE))

#Set working directory (assumes data file is available in same location as code)
current_path = rstudioapi::getActiveDocumentContext()$path 
setwd(dirname(current_path))

#Read Data File
AHDB_Exp1_BloodData <- read.csv("AHDB_Exp1_BloodData.csv")
str(AHDB_Exp1_BloodData)

# Define Plot Theme
plot_theme <- theme_classic(base_size = 18) +
  theme(axis.title.y = element_text(vjust = 1.5), 
        axis.title.x = element_blank()) +
  theme(plot.margin = unit(c(.3, .3, .6, .6), "cm"), 
        line = element_line(linewidth = 1.25)) + 
  theme(axis.line.x = element_line(colour = "black", linewidth = 1.25),
        axis.line.y = element_line(colour = "black", linewidth = 1.25)) + 
  theme(axis.text.x = element_text(angle = 360, hjust = .5, 
                                   margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"), 
                                   color = "black"),
        axis.text.y = element_text(margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"), 
                                   color = "black")) +
  theme(axis.ticks.length = unit(-0.3, "cm"),
        axis.ticks = element_line(color = "black", linewidth = 1.25)) + 
  theme(panel.spacing = unit(2, "lines")) +
  theme(strip.background = element_blank()) +
  #theme(legend.position = "none") + 
  theme(plot.title = element_blank()) 
```

# Data Processing & Filtering 
```{r}

# Filter AHDB_Exp1_BloodData to baseline timepoint only
# This is a post hoc analysis of heat-stress induced mortality, therefore treatment A (control) was removed due to individuals not experiencing the heat treatment

AHDB_Exp1_Baseline <- AHDB_Exp1_BloodData %>%
  filter(TimePoint == "Baseline",
         Treatment     != "A")

#Convert Died in Treatment to factor
AHDB_Exp1_Baseline <- AHDB_Exp1_Baseline %>%
  mutate(
    Died_InTrt = factor(Died_InTrt,
                        levels = c("No", "Yes")))  


# Define columns to convert to numeric 

cols_to_convert <- c("Glucose_1", "Glucose_2", "Glucose_3",
                     "Ketone_1", "Ketone_2", "Ketone_3", "Hematocrit")

# Replace "No" with NA in the specified columns
AHDB_Exp1_Baseline[cols_to_convert] <- lapply(AHDB_Exp1_Baseline[cols_to_convert], function(x) {
  x[x == "No"] <- NA
  x
})

# Convert these columns to numeric
AHDB_Exp1_Baseline[cols_to_convert] <- lapply(AHDB_Exp1_Baseline[cols_to_convert], as.numeric)

# Calculate average values for Glucose and Ketones
AHDB_Exp1_Baseline$Glucose_Avg <- rowMeans(AHDB_Exp1_Baseline[, c("Glucose_1", "Glucose_2", "Glucose_3")],
                                                               na.rm = TRUE)
AHDB_Exp1_Baseline$Ketone_Avg <- rowMeans(AHDB_Exp1_Baseline[, c("Ketone_1", "Ketone_2", "Ketone_3")],
                                                              na.rm = TRUE)

# Create a Female ZEFI baseline dataset 
AHDB_Baseline_Female <- AHDB_Exp1_Baseline %>%
  filter(
    Sex           == "F",
    TimePoint     == "Baseline"
  )

# Create a Younger Female ZEFI baseline dataset 
AHDB_Baseline_YoungerFemale <- AHDB_Baseline_Female %>%
  filter(
    Sex           == "F",
    AgeCategory     == "Younger"
  )

```

# Data Analysis 
# Hypothesis 1 . Females are more likely to die from acute heat (~ 43 – 43.5C for 5 hours) than males.( Includes Mom_ID as a random effect) 

# Calculate Sample Size by Sex (Does not account for age)

```{r}

AHDB_Exp1_Baseline %>%
  distinct(ID_Band, Sex) %>%
  summarise(
    Total   = n(),
    Males   = sum(Sex == "M"),
    Females = sum(Sex == "F")
  )

```
# Firth Model -> Sex 

```{r}

# This runs Firth's logistic regression with the penalized likelihood approach)
# Useful for data with reduced sample size and complete separation in variables (DeathInTrt - No Males Died)

firth_fit_sex <- logistf(Died_InTrt ~  Sex ,  
                         data = AHDB_Exp1_Baseline)
summary(firth_fit_sex)

exp(coef(firth_fit_sex)) 
exp(confint(firth_fit_sex))

```

# Calculate pseudo-R2 (Proportion of variation of independent variable explained by the dependent variable)
```{r}

ll.null <- firth_fit_sex$loglik[1]
ll.proposed <- firth_fit_sex$loglik[2]
pseudoR2 <- (ll.proposed - ll.null) / ll.null
cat("Pseudo R²:", pseudoR2, "\n")

# Result: 5.9%

```

# Predicted probabilities for each Sex

```{r}
death_data <- data.frame(Sex = c("M", "F"))
predicted_prob_sex <- predict(firth_fit_sex, death_data, type = "response")
prob_table <- data.frame(Sex = death_data$Sex, Predicted_Probability = predicted_prob_sex)
print(prob_table)

# Results: Females have a 16.7 % likelihood of succumbing to acute heat exposure compared to 2.5% for the Males. 

```

# Plot predicted probabilities

```{r}
predicted_prob_sex_plot <-ggplot(death_data, aes(x = Sex, y = predicted_prob_sex, fill = Sex)) +
  geom_point(aes(color = Sex), size = 5) +
  xlab("Sex") +
  ylab("Predicted Probability of Death in Acute Heat") +
  plot_theme

predicted_prob_sex_plot

ggsave(
  "C:/Users/aliam/Box/NSF_Damage-Fitness project-Research/Experiment_2024Fall_AcuteHeat_DamageBiomarkers_Exp1/Data/DataAnalyses/Death_AA/Figures/MS1_Predicted_Death.jpg",
  predicted_prob_sex_plot,
  dpi    = 250,
  width  = 10,
  height = 8
)
```

# Calculate observed death proportions 

```{r}
obs_death_sex <- AHDB_Exp1_Baseline %>%
  mutate(Death = if_else(Died_InTrt == "Yes", 1, 0)) %>%
  group_by(Sex) %>%
  summarize(
    N = n(),
    DeathCount = sum(Death, na.rm = TRUE),
    Proportion = DeathCount / N
  )
print(obs_death_sex)

# Observed Female Death Proportion: 0.162
# Observed Male Death Proportion: 0

```
# Observed vs Predicted Prob of Death Plot

```{r}
pred_obs_sex <- ggplot() +
  # Plot observed death proportions (blue circles)
  geom_point(data = obs_death_sex, 
             aes(x = Sex, y = Proportion, color = "Observed"), 
             size = 4) +
  # Plot model predictions (red triangles)
  geom_point(data = prob_table, 
             aes(x = Sex, y = Predicted_Probability, color = "Predicted"), 
             size = 4, shape = 17) +
  scale_color_manual(values = c("Observed" = "blue", "Predicted" = "red"),
                     name = "Legend",
                     labels = c("Observed", "Predicted")) +
  labs(title = "Observed vs. Predicted Death Proportions by Sex",
       x = "Sex",
       y = "Death Proportion") +
  ylim(0, 1) +
  plot_theme

pred_obs_sex
# Model predictions in line with observed death data

ggsave(
  "C:/Users/aliam/Box/NSF_Damage-Fitness project-Research/Experiment_2024Fall_AcuteHeat_DamageBiomarkers_Exp1/Data/DataAnalyses/Death_AA/Figures/MS1_ObservedvsPredicted_Death.jpg",
  predicted_prob_sex_plot,
  dpi    = 250,
  width  = 10,
  height = 8
)

```

# Calculate Sample Size by Age & Sex for Firth Model that includes both sex and age

```{r}

AHDB_Exp1_Baseline %>%
  distinct(ID_Band, Sex, AgeCategory) %>%
  group_by(Sex, AgeCategory) %>%
  summarise(
    Count = n(),
    .groups = "drop"
  )

```

# Firth Model - Sex + Age 
```{r}

firth_fit_sex_age<- logistf(Died_InTrt ~  Sex + AgeCategory, #+ Sex*AgeCategory,  
                            data = AHDB_Exp1_Baseline)
summary(firth_fit_sex_age)
#Age & Sex interaction was removed by backward regression 

exp(coef(firth_fit_sex_age)) 
exp(confint(firth_fit_sex_age))
```

# Calculate pseudo-R2 (Proportion of variation of independent variable explained by the dependent variable)

```{r}

ll.null <- firth_fit_sex_age$loglik[1]
ll.proposed <- firth_fit_sex_age$loglik[2]
pseudoR2 <- (ll.proposed - ll.null) / ll.null
cat("Pseudo R²:", pseudoR2, "\n")


```
# Calculate Predicted Probabilities with both Age & Sex

```{r}

death_data <- data.frame(
  Sex         = c("M","F","F","M"),
  AgeCategory = c("Younger","Older","Younger","Older")
)
predicted_prob_sex_age <- predict(firth_fit_sex_age, death_data, type = "response")
prob_table_sex_age <- data.frame(Sex = death_data$Sex, AgeCategory =death_data$AgeCategory, Predicted_Probability = predicted_prob_sex_age)
print(prob_table_sex_age)

# Results: Younger Females have a 22.8% likelihood of succumbing to acute heat exposure compared to 3.1% for the younger males whereas older females have 6.3% chance of mortality compared to 0.7% for older males. 


```
# Hypothesis 2. Younger adult females are more likely to diet from acute heat (~ 43–43.5C for 5 hours) than older young adult females. (Includes Mom_ID as a random effect)

# Firth - Younger Females Only

```{r}
firth_fit_age <- logistf(Died_InTrt ~  AgeCategory, 
                         data = AHDB_Baseline_Female)
summary(firth_fit_age)
exp(coef(firth_fit_age)) 
exp(confint(firth_fit_age))

```
# Calculate "Pseudo R²
```{r}

ll.null <- firth_fit_age$loglik[1]
ll.proposed <- firth_fit_age$loglik[2]
pseudoR2 <- (ll.proposed - ll.null) / ll.null
cat("Pseudo R²:", pseudoR2, "\n")

```
# Predicted probabilities for each AgeCategory
```{r}

deathdata_H2 <- data.frame(AgeCategory = c("Older", "Younger"))
predicted_prob_age <- predict(firth_fit_age, newdata = deathdata_H2, type = "response")
prob_table <- data.frame(Age = deathdata_H2$Age, Predicted_Probability = predicted_prob_age)
print(prob_table)

```
# Plot predicted probabilities

```{r}

predicted_prob_age_plot <-ggplot(deathdata_H2, aes(x = AgeCategory, y = predicted_prob_age, fill = AgeCategory)) +
  geom_point(aes(color = AgeCategory), size = 5) +
  xlab("Age") +
  ylab("Predicted Probability of Death in Acute Heat") +
  plot_theme

predicted_prob_age_plot

ggsave(
  "C:/Users/aliam/Box/NSF_Damage-Fitness project-Research/Experiment_2024Fall_AcuteHeat_DamageBiomarkers_Exp1/Data/DataAnalyses/Death_AA/Figures/MS1_FemalePredicted_AgeDeath.jpg",
  predicted_prob_age_plot,
  dpi    = 250,
  width  = 10,
  height = 8
)

```

# Calculate observed death probabilites for younger females

```{r}

obs_death_age <- AHDB_Baseline_Female %>%
  mutate(Death = if_else(Died_InTrt == "Yes", 1, 0)) %>%
  group_by(AgeCategory) %>%
  summarize(
    N = n(),
    DeathCount = sum(Death, na.rm = TRUE),
    Proportion = DeathCount / N
  )
print(obs_death_age)

```

#  Observed vs Predicted Prob of Death Plot

```{r}

obs_death_age <- ggplot() +
  # Plot observed death proportions (blue circles)
  geom_point(data = obs_death_age, 
             aes(x = AgeCategory, y = Proportion, color = "Observed"), 
             size = 4) +
  # Plot model predictions (red triangles)
  geom_point(data = prob_table, 
             aes(x = Age, y = Predicted_Probability, color = "Predicted"), 
             size = 4, shape = 17) +
  scale_color_manual(values = c("Observed" = "blue", "Predicted" = "red"),
                     name = "Legend",
                     labels = c("Observed", "Predicted")) +
  labs(title = "Observed vs. Predicted Death Proportions by Age",
       x = "Age",
       y = "Death Proportion") +
  ylim(0, 1) +
  plot_theme

obs_death_age
#Model predictions in line with observed death data

ggsave(
  "C:/Users/aliam/Box/NSF_Damage-Fitness project-Research/Experiment_2024Fall_AcuteHeat_DamageBiomarkers_Exp1/Data/DataAnalyses/Death_AA/Figures/MS1_FemaleObservedvPredicted_AgeDeath.jpg",
  obs_death_age,
  dpi    = 250,
  width  = 10,
  height = 8
)

```

# Hypothesis 3. Among young adult females, there will be metabolic or damage variables in the baseline blood samples (2 weeks prior) may correlated with probability of dying due to acute heat (~43 – 43.5C for 5 hours). (Includes Mom_ID as a random effect)

# Glucose Model -> To see whether baseline glucose levels predict death in younger females (Includes Individuals with lysed plasma)

# Remove rows with missing values
```{r}

Younger_females_Glu_lysed <- AHDB_Baseline_YoungerFemale %>%
  filter(!is.na(Glucose_Avg))

```

# Calculate Mean Glucose Values & plot residuals
```{r}

mean(Younger_females_Glu_lysed$Glucose_Avg)
sd(Younger_females_Glu_lysed$Glucose_Avg)
min(Younger_females_Glu_lysed$Glucose_Avg)
max(Younger_females_Glu_lysed$Glucose_Avg)
hist(Younger_females_Glu_lysed$Glucose_Avg)

```
# Calcualte Sample size For Glucose Model
```{r}

Younger_females_Glu_lysed %>%
  distinct(ID_Band, Sex) %>%
  summarise(
    Total   = n(),
    Females = sum(Sex == "F")
  )

```
# GLMER -> Glucose Model (Includes Lysed Plasma Individuals)

```{r}

glmer_model_Glu <- glmer(Died_InTrt ~  Glucose_Avg +
                           (1 | Mom_ID) + (1 | Cohort),
                         data = Younger_females_Glu_lysed,
                         family = binomial)

summary(glmer_model_Glu)

#Not significant
```
# Remove rows with plasma color other than (“U” or “U/L”, "Y") and calculate how many individuals were removed
# Logic: Hemolysis could release compounds (NADH, Catalase) that may interfere and skew glucose and ketone readings

```{r}

# Remove rows with missing values
Younger_females_Glu_NotLysed <- AHDB_Baseline_YoungerFemale %>%
  filter(!is.na(Glucose_Avg))

# Capture dataset before plasma‐color filtering
Younger_females_Glu_before_plasma <- Younger_females_Glu_NotLysed

# Remove individuals with lysed plasma appearance
Younger_females_Glu_NotLysed <- Younger_females_Glu_NotLysed %>%
  filter(Plasma_Color %in% c("U", "U/L", "Y"))

#Counts how many individuals were removed due to lysis or being yolky
removed_plasma_rows <- Younger_females_Glu_before_plasma %>%
  filter(!Plasma_Color %in% c("U", "U/L", "Y"))

# Total rows (samples) removed
n_removed_rows <- nrow(removed_plasma_rows)
cat("Rows removed due to plasma appearance:", n_removed_rows, "\n")

# Unique individuals removed (by ID_Band)
n_removed_individuals <- removed_plasma_rows %>%
  distinct(ID_Band) %>%
  nrow()
cat("Unique individuals removed:", n_removed_individuals, "\n")
# 13 individuals removed due to plasma appearance

```
# Recalculate Sample Size

```{r}

Younger_females_Glu_NotLysed %>%
  distinct(ID_Band, Sex) %>%
  summarise(
    Total   = n(),
    Females = sum(Sex == "F")
  )

```

# Glucose Model (Excludes individuals with lysed plasma)

```{r}
glmer_model_Glu <- glmer(Died_InTrt ~  Glucose_Avg +
                           (1 | Mom_ID) + (1 | Cohort),
                         data = Younger_females_Glu_NotLysed,
                         family = binomial)

summary(glmer_model_Glu)

# Not significant, high variance in Mom_ID. No cohort effect observed. 
```
# GLMER -> Ketone Model (Includes Lysed Plasma Individuals) -> To see whether baseline ketone levels predict death in younger females due to acute heat

# Remove rows with missing values
```{r}

Younger_females_Ket_Lysed <- AHDB_Baseline_YoungerFemale %>%
  filter(!is.na(Ketone_Avg))

```

# Calculate Mean Ketone Values & plot residual

```{r}

mean(Younger_females_Ket_Lysed$Ketone_Avg)
sd(Younger_females_Ket_Lysed$Ketone_Avg)
min(Younger_females_Ket_Lysed$Ketone_Avg)
max(Younger_females_Ket_Lysed$Ketone_Avg)
hist(Younger_females_Ket_Lysed$Ketone_Avg)

```

# Calcualte Sample size For Ketone Model (Includes Lysed Plasma Appearance Individuals)
```{r}

Younger_females_Ket_Lysed %>%
  distinct(ID_Band, Sex) %>%
  summarise(
    Total   = n(),
    Females = sum(Sex == "F")
  )

# Sample SIze: 43
```
# GLMER -> Ketone Model (Includes Lysed Plasma Individuals)

```{r}

glmer_model_Ket <- glmer(Died_InTrt ~  Ketone_Avg +
                           (1 | Mom_ID) + (1 | Cohort),
                         data = Younger_females_Ket_Lysed,
                         family = binomial)

summary(glmer_model_Ket)

#Not significant
```
# Remove rows with plasma color other than (“U” or “U/L”, "Y") and calculate how many individuals were removed
# Logic: Hemolysis could release compounds (NADH, Catalase) that may interfere and skew glucose and ketone readings

```{r}

# Remove rows with missing values
Younger_females_Ket_NotLysed <- AHDB_Baseline_YoungerFemale %>%
  filter(!is.na(Ketone_Avg))

# Capture dataset before plasma‐color filtering
Younger_females_Ket_before_plasma <- Younger_females_Ket_NotLysed

# Remove individuals with lysed plasma appearance
Younger_females_Ket_NotLysed <- Younger_females_Ket_NotLysed %>%
  filter(Plasma_Color %in% c("U", "U/L", "Y"))

#Counts how many individuals were removed due to lysis or being yolky
removed_plasma_rows_ket <- Younger_females_Ket_before_plasma %>%
  filter(!Plasma_Color %in% c("U", "U/L", "Y"))

# Total rows (samples) removed
n_removed_rows_ket <- nrow(removed_plasma_rows_ket)
cat("Rows removed due to plasma appearance:", n_removed_rows_ket, "\n")

# Unique individuals removed (by ID_Band)
n_removed_individuals_ket <- removed_plasma_rows_ket %>%
  distinct(ID_Band) %>%
  nrow()
cat("Unique individuals removed:", n_removed_individuals_ket, "\n")
# 12 individuals removed due to plasma appearance

```
# Recalculate Sample Size

```{r}

Younger_females_Ket_NotLysed %>%
  distinct(ID_Band, Sex) %>%
  summarise(
    Total   = n(),
    Females = sum(Sex == "F")
  )

# Sample Size: 31
```

# Ketone Model (Excludes individuals with lysed plasma)

```{r}

glmer_model_Ket <- glmer(Died_InTrt ~  Ketone_Avg +
                           (1 | Mom_ID) + (1 | Cohort),
                         data = Younger_females_Ket_NotLysed,
                         family = binomial)

summary(glmer_model_Ket)

# Ketones is significant, high variance in Mom_ID. No cohort effect observed. 
```
# Hematocrit Model ## To see whether baseline hematocrit levels predict death in  younger females due to acute heat. 

# Remove rows with missing values
```{r}

Younger_females_Hem <- AHDB_Baseline_YoungerFemale %>%
  filter(!is.na(Hematocrit))

```

# Calculate Mean Hematocrit Values & plot residuals
```{r}

mean(Younger_females_Hem$Hematocrit)
sd(Younger_females_Hem$Hematocrit)
min(Younger_females_Hem$Hematocrit)
max(Younger_females_Hem$Hematocrit)
hist(Younger_females_Hem$Hematocrit)

```

#Create log-transformed Hematocrit columns
```{r}

Younger_females_Hem$logHematocrit <- log10(Younger_females_Hem$Hematocrit)
hist(Younger_females_Hem$logHematocrit)

#Better

```
# Calculate Sample Size

```{r}

Younger_females_Hem %>%
  distinct(ID_Band, Sex) %>%
  summarise(
    Total   = n(),
    Females = sum(Sex == "F")
  )

# Sample Size: 40
```

# GLMER -> Hematocrit Model (Includes Lysed Individuals)

```{r}

glmer_model_Hematocrit <- glmer(Died_InTrt ~  logHematocrit +
                                  (1 | Mom_ID) + (1 | Cohort),
                                data = Younger_females_Hem,
                                family = binomial)

summary(glmer_model_Hematocrit)

# Hematocrit not significant - low variance in mom ID, no cohort effect observed 
```

# GLMER -> Hematocrit Model (Excludes Lysed Individuals)

# Remove rows with plasma color other than (“U” or “U/L”, "Y") and calculate how many individuals were removed
# Logic: Hemolysis could release compounds (NADH, Catalase) that may interfere and skew glucose and ketone readings

```{r}

# Remove rows with missing values
Younger_females_Hem_NotLysed <- AHDB_Baseline_YoungerFemale %>%
  filter(!is.na(Hematocrit))

# Capture dataset before plasma‐color filtering
Younger_females_Hem_before_plasma <- Younger_females_Hem_NotLysed

# Remove individuals with lysed plasma appearance
Younger_females_Hem_NotLysed <- Younger_females_Hem_NotLysed %>%
  filter(Plasma_Color %in% c("U", "U/L", "Y"))

#Counts how many individuals were removed due to lysis or being yolky
removed_plasma_rows <- Younger_females_Hem_before_plasma %>%
  filter(!Plasma_Color %in% c("U", "U/L", "Y"))

# Total rows (samples) removed
n_removed_rows <- nrow(removed_plasma_rows)
cat("Rows removed due to plasma appearance:", n_removed_rows, "\n")

# Unique individuals removed (by ID_Band)
n_removed_individuals <- removed_plasma_rows %>%
  distinct(ID_Band) %>%
  nrow()
cat("Unique individuals removed:", n_removed_individuals, "\n")
# 10 individuals removed due to plasma appearance

```
# Recalculate Sample Size

```{r}

Younger_females_Hem_NotLysed %>%
  distinct(ID_Band, Sex) %>%
  summarise(
    Total   = n(),
    Females = sum(Sex == "F")
  )

# Sample Size: 30
```

#Create log-transformed Hematocrit columns
```{r}

Younger_females_Hem_NotLysed$logHematocrit <- log10(Younger_females_Hem_NotLysed$Hematocrit)
hist(Younger_females_Hem_NotLysed$logHematocrit)

```
# GLMER -> Hematocrit Model (Includes Lysed Individuals)

```{r}

glmer_model_Hematocrit <- glmer(Died_InTrt ~  logHematocrit +
                                  (1 | Mom_ID) + (1 | Cohort),
                                data = Younger_females_Hem_NotLysed,
                                family = binomial)

summary(glmer_model_Hematocrit)
# Not significant, low variance for cohort and mom_id
```
# ALL SUBSEQUENT ANALYSES NOT REMOVE MOM_ID AS A RANDOM EFFECT 

# GLMER -> Glucose Model (WITH LYSED BUT NO Mom_ID)

```{r}

glmer_model_Glu <- glmer(Died_InTrt ~  Glucose_Avg +
                            (1 | Cohort),
                         data = Younger_females_Glu_lysed,
                         family = binomial)

summary(glmer_model_Glu)


# No change in results
```
# GLMER -> Glucose Model (WITHOUT LYSED INDIVIDUALS & Mom_ID)

```{r}

glmer_model_Glu <- glmer(Died_InTrt ~  Glucose_Avg +
                            (1 | Cohort),
                         data = Younger_females_Glu_NotLysed,
                         family = binomial)

summary(glmer_model_Glu)

# No change in results

```


# GLMER -> Ketone Model (WITH LYSED BUT NO Mom_ID)

```{r}

glmer_model_Ket <- glmer(Died_InTrt ~  Ketone_Avg +
                            (1 | Cohort),
                         data = Younger_females_Ket_Lysed,
                         family = binomial)

summary(glmer_model_Ket)


# No change in results
```
# GLMER -> Ketone Model (WITHOUT LYSED INDIVIDUALS & Mom_ID)

```{r}

glmer_model_Ket <- glmer(Died_InTrt ~  Ketone_Avg +
                            (1 | Cohort),
                         data = Younger_females_Ket_NotLysed,
                         family = binomial)

summary(glmer_model_Ket)


# Ketones are no longer significant (p-value = 0.749)

```
# GLMER -> Hematocrit Model (WITH LYSED BUT NO Mom_ID)

```{r}

glmer_model_Hem <- glmer(Died_InTrt ~  logHematocrit +
                            (1 | Cohort),
                         data = Younger_females_Hem,
                         family = binomial)

summary(glmer_model_Hem)

# Convergence Error at Maximum 

glmer_model_Hem <- glmer(Died_InTrt ~  Hematocrit +
                            (1 | Cohort),
                         data = Younger_females_Hem,
                         family = binomial)

summary(glmer_model_Hem)
# Scaling Error 

# Scale Predictor
Younger_females_Hem <- Younger_females_Hem %>%
  mutate(
    Hematocrit_s = as.numeric(scale(Hematocrit))
  )

glmer_model_Hem <- glmer(Died_InTrt ~  Hematocrit_s +
                            (1 | Cohort),
                         data = Younger_females_Hem,
                         family = binomial)

summary(glmer_model_Hem)

# Results are same as before i.e. hematocrit is not significant
```
# GLMER -> Hematocrit Model (WITHOUT LYSED INDIVIDUALS & Mom_ID)

```{r}

glmer_model_Hem <- glmer(Died_InTrt ~  logHematocrit +
                            (1 | Cohort),
                         data = Younger_females_Hem_NotLysed,
                         family = binomial)

summary(glmer_model_Hem)

# Same results as before, hematocrit is not significant (0.220)


```
